---
title: "Fitting Distributions to Data"
author: "Stat 241"
output:
  html_document:
    theme: yeti
    css: ../styles/ds303-notes.css
  df_document: default
  pdf_document: default
---

\def\tor{\operatorname{or}}
\def\tand{\operatorname{and}}
\def\tnot{\operatorname{not}}
\def\Prob{\operatorname{P}}
\def\E{\operatorname{E}}
\def\Var{\operatorname{Var}}
\def\SD{\operatorname{SD}}
\def\Binom{{\sf Binom}}
\def\Norm{{\sf Norm}}
\def\Chisq{{\sf Chisq}}
\def\Unif{{\sf Unif}}
\def\Exp{{\sf Exp}}
\def\Beta{{\sf Beta}}
\def\Gamm{{\sf Gamma}}
\def\Weibull{{\sf Weibull}}
\def\Pois{{\sf Pois}}

<script src="../styles/folding.js"></script>

```{r hooks, message=FALSE, include = FALSE}
knitr::knit_hooks$set(fold = function(before, options, envir) {
  if (before) {
    return(glue::glue('<div fold = "{options$fold}">'))
  } else {
    return('</div>\n')
  }
})
```


```{r setup, include=FALSE}
onLine <- TRUE
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 5, 
  fig.height = 2,
  fig.align = "center"
)
library(mosaic)
library(pander)
library(mosaicCalc)
library(triangle)
theme_set(theme_bw())
set.seed(20210108)
options(digits = 3)
```
 
<style>
div.boxed {
  border-style: solid;
  border-color: navy;
  border-width: 2px;
  border-radius: 10px;
  padding 20px;
</style>



**Problem:** 
Suppose we have data for a continuous random variable X.  Can we find (estimate) the distribution for X?

* Part 1:  What type of distribution?  Exponential, normal, uniform, gamma, beta, \dots 
* Part 2:  What are the values of the parameters for the distribution?

**Three Methods**

1. Method of moments 

    * Only requires summaries of data, not complete data.
    
    * Can be done by hand in many important cases
    
    * But there are better ways to get estimates (better statistical properties)
    

2. Maximum likelihood 

    * Harder to do by hand, but better estimates
    
    * Requires the full data set, lots of computation, and optimization

    * This is what `fitdistr()` and `gf_fitdistr()` do)
    
3. Kernel density estimate 

    * Let's us skip step 1 
    
    * This is what `gf_dens()` and `gf_density()`
    
    * Big idea: replace each data point with a little curve (a kernel) and add them together.
    
    * Get different results depending on the choice of curve.

All of these methods work better with large data sets than with small data sets --
it is hard to determine the overall shape of a distribution without a large data set.

## Method of Moments

Big idea: 

1. Express the mean, variance, etc. of the distribution in terms of the parameters of 
the distribution.  (You can look this information up in our book or on Wikipedia.)

2. Set them equal to the mean, variance, etc. of your data set.

3. Solve.


**Example:** A sample has mean 15 and standard deviation 3. Use the method of moments to 
fit a uniform distribution.

<div class="explain" label = "Show solution">
-----------------------------------------
 moment        data       distribution         
------------ -------- -------------------
  mean          15        $\frac{a+b}{2}$
  
 variance       3^2       $\frac{(b-a)^2}{12}$
-----------------------------------------

So we need to solve this system of equations:

\begin{align*}
15 &= \frac{a+b}{2} \\
3^2  &= \frac{(b-a)^2}{12} \\
\end{align*}

If we let $m = \frac{a + b}{2}$ and let $d = b-a$, our equations become 
  
\begin{align*}
15 &= m \\
3^2  &= \frac{d^2}{12} \\
\end{align*}

This is much easier.  We see that $d = \sqrt{9 \cdot 12}$. Let let R do the arithmetic.

```{r}
m <- 15; m
d <- sqrt(9*12); d 
a <- m - d/2; a
b <- m + d/2; b
```

So $\hat a = 9.8$, $\hat b = 20.2$.

</div>
<br>
		 		 
**Example:** A sample has mean 15 and standard deviation 3. Use the method of moments to 
fit a normal distribution.

<div class="explain" label = "Show solution">
This is a very easy case.

--------------------------------------------------
 moment                 data       distribution         
--------------------- -------- -------------------
  mean                   15        $\mu$
  
 standard deviation       3        $\sigma$
--------------------------------------------------

So we get $\hat \mu = 15$ and $\hat \sigma = 3$.
</div>
<br>

**Example:** A sample has mean 15 and standard deviation 3. Use the method of moments to 
fit an exponential distribution.

<div class="explain" label = "Show solution">

In this case, we only need the mean.
-----------------------------------
 moment     data     distribution         
--------- -------- ----------------
  mean       15      $1/ \lambda$
-----------------------------------

So $\hat \lambda = 1/15$.
</div>
<br>

**Example:** A sample has mean 15 and standard deviation 3. Use the method of moments to 
fit an gamma distribution.

<br>

**General Example**: A sample has mean = ________ and standard deviation = _________, 
use the method of moments to fit a ________ distribution.

<br>


* Might be easy or hard depending on the algebra involved.

* If there is only one parameter, we only need to use the mean.

* There are ways to use distributions with more than 2 parameters, but we won't cover that in this class.

* **No guarantee that the fit is a good fit**.  You can fit any distribution using
any method to any data. Other methods (plots, for example) need to be used to see whether
the fit is any good.

## Maximum Likelihood (`fitdistr()`)

**Example:  Wind speed**

<div class="explain" label="Show example">

We can use the following code to load a data set that contains three year's worth 
of mean hourly wind speeds (mph) in Twin Falls, ID.  This kind of data is often used
to estimate how much power could be generated from a windmill placed in a given location.

```{r include = !onLine,eval = !onLine}
Wind <- read.csv("data/stob/TwinfallsWind.csv")
head(Wind, 2)
tail(Wind, 2)
histogram( ~ speed, data = Wind, binwidth = 1 )
```

```{r include = onLine,eval = onLine}
Wind <- 
  read.csv("https://rpruim.github.io/Engineering-Statistics/data/stob/TwinfallsWind.csv")
head(Wind, 2)
tail(Wind, 2)
gf_histogram( ~ speed, data = Wind, binwidth = 1 )
```

As we can see, the distribution is skewed, but it doesn't look like an 
exponential distribution would be a good fit.  Of the distributions we have seen,
it seems like a Weibull or Gamma distribution would be a potentially good choice.
A Weibull model has often been used as a model for mean hourly wind speed, and the 
shape of our histogram indicates that this is a reasonable family of distributions.

Which Weibull distribution is the best model for our data?

```{r fitdistr-windspeed, error = TRUE}
fitdistr( Wind$speed, "weibull" )
```

For `fitdistr()()` to fit a Weibull distribution, all of the data must be positive,
but our data includes some 0's.  
```{r zeros}
tally( ~ (speed == 0), data = Wind)
```

Let's see how small the smallest non-zero measurements 
are.

```{r}
min( ~ speed, data = Wind %>% filter(speed > 0))
```

This may well be a simple rounding issue, since the wind speeds are recorded to the 
nearest 0.01 and 0.01 is the smallest positive value.
Let's create a new variable that moves each value of 0 to 0.0025 and try again.
Why 0.0025?  If we think that 0.01 represents anything in the range 0.005 to 0.015, which
would round to 0.01, then 0 represents anything in the range 0 to 0.005.  
0.0025 is the middle of that range.

```{r fitdistr-windspeed2, warning = FALSE}
Wind <- Wind %>% mutate(speed2 = ifelse( speed > 0, speed, 0.0025))
fitdistr( Wind$speed2, "weibull" )
```

```{r include = FALSE}
mle <- fitdistr( Wind$speed2, "weibull" )
```

This says that the best fitting (in the sense of maximum likelihood) Weibull
distribution is the $\Weibull(`r paste(round(mle$est,2),sep = ",")`)$-distribution.

Let's see how well it fits:

```{r xhistogram-Wind,warning = FALSE}
gf_dhistogram( ~ speed2, data = Wind) %>%
  gf_fitdistr( ~ speed2, data = Wind, dist = "weibull")
```

This can be abbreviated a bit:

```{r xhistogram-Wind-2,warning = FALSE}
gf_dhistogram( ~ speed2, data = Wind) %>%
  gf_fitdistr(dist = "weibull")
```

</div>


<!-- **Example:** -->
<!-- The variable `avg_drinks` in the data frame HELPrct contains data for the -->
<!-- average number of alcoholic drinks consumed per day for each participant.  -->
<!-- What distribution fits these data well? -->

<!-- <div class = "explain" label = "Show solution"> -->

<!-- Note: the distribution is different for the different substance groups. -->

<!-- ```{r} -->
<!-- gf_dens( ~ avg_drinks, data = HELPrct, color = ~ substance) -->
<!-- ``` -->

<!-- So we should probably fit separately to each group.  Let's try the heroin group. -->

<!-- ```{r} -->
<!-- Heroin <- HELPrct %>% filter(substance == "heroin") -->
<!-- gf_dhistogram(~ avg_drinks, data = Heroin) -->
<!-- ``` -->

<!-- That is a skewed distribution, so we should try something like exponential or gamma or Weibull. -->

<!-- ```{r} -->
<!-- fitdistr(Heroin$avg_drinks, "exponential") -->
<!-- fitdistr(Heroin$avg_drinks, "gamma", start = list(shape = 2, rate = 5)) -->

<!-- ``` -->

## Some Practice

**1.** Suppose a data set has a sample mean of 6 and a standard deviation of 2. 
Use the method of moments to compute estimates of the shape and rate paramters
for a gamma distribution.

**2.** Fit a gamma distribution to the wind speed data. Use both the method of moments
and `fitdistr()`. How do the gamma and Weibull fits compare?

**3.** The `weight` variable in the `ChickWeight` data frame contains the weights (g) 
of baby chicks  at different times under different feeds.

a.	Plot a histogram of the weights.
 
b.	Of the distributions:  uniform, exponential, normal, and gamma, which look like they might fit 
the data well?  

<!-- Either Exponential or gamma is a good answer -->

c.	Use `fitdistr()` to fit the type of distribution in (b) to the data.  What is the estimate of the parameter(s) for the distribution?  

<!-- Estimated value is rate = .0082 if exponential -->
<!-- Estimated values are shape = 3.15, rate = .026 if gamma -->

d.	Plot the histogram with the fitted density function.
  
e.	Use `fitdistr()` to fit one of the other types of distributions from (b) 
and plot the histogram with the density curve for that type of distribution. 
How does the fit look? 


**4.**	This problem use the `Womenlf` data set from the `car` package.
The `hincome` variable contains household income for 263 Canadian women.

a. Plot a histogram of these incomes.

b. Use the method of moments to fit a normal distribution to these data.

c. Use `fitdistr()` to fit a normal distribution to these data. How does this compare to (b)?

d. Plot the histogram with the fitted normal density function.

e. Now fit a gamma distribution to this data. 
What are the estimates for the shape and rate of the fitted gamma distribution?

f. Plot the histogram with the fitted gamma density function.
 
g. Compare the quality of the fits with normal vs gamma.  Does one clearly fit
better than the other? 

<!-- They both seem to fit reasonably well.  Given that the data is skewed a bit to -->
<!-- the right, maybe the gamma fits better. -->

**5.** The dataframe `morley` contains 100 measurements that Morley made of the speed
of light.  These values are contained in the column `Speed`.

a. Plot the histogram of this data.
 
b. Which kind of distribution best fits the data? 

c. Find the estimated parameters for this distribution.

d. Plot the histogram with the fitted density curve.  



