---
title: "Expected Value and Variance of Random Variables"
author: "Stat 241"
output:
  html_document:
    theme: yeti
    css: ../styles/ds303-notes.css
  df_document: default
  pdf_document: default
---

\def\tor{\operatorname{or}}
\def\tand{\operatorname{and}}
\def\tnot{\operatorname{not}}
\def\Prob{\operatorname{P}}
\def\E{\operatorname{E}}
\def\Var{\operatorname{Var}}
\def\SD{\operatorname{SD}}
\def\Binom{{\sf Binom}}
\def\Norm{{\sf Norm}}
\def\Unif{{\sf Unif}}

<script src="../styles/folding.js"></script>

```{r hooks, message=FALSE, include = FALSE}
knitr::knit_hooks$set(fold = function(before, options, envir) {
  if (before) {
    return(glue::glue('<div fold = "{options$fold}">'))
  } else {
    return('</div>\n')
  }
})
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 5, 
  fig.height = 2,
  fig.align = "center"
)
library(mosaic)
library(pander)
library(mosaicCalc)
library(triangle)
theme_set(theme_bw())
set.seed(20210108)
```

## Warm-Up

**Example 1.**
Suppose a student has taken $10$ courses
and received $5$ A's, $4$ B's, and $1$ C.  Using the traditional numerical scale where 
an A is worth $4$, a B is worth $3$, and a C is worth $2$, what is this student's 
GPA (grade point average)?

\vfill
\vfill

<div class = "explain" label = "Show solution">
The first thing to notice is that $\frac{4 + 3 + 2}{3} = 3$ is \emph{not} correct.
  We cannot simply add up the values and divide by the number of values.  Clearly
  this student should have a GPA that is higher than $3.0$, since there were more A's than
  C's.

Consider now a correct way to do this calculation:
\begin{align*}
	\mbox{GPA} &= \frac{4 + 4 + 4 + 4 + 4 + 3 + 3 + 3 + 3 + 2}{10}
	\\[2mm]
	& = \frac{5\cdot 4 + 4\cdot 3 + 1 \cdot 2}{10} \\[1mm]
	& = \frac{5}{10} \cdot 4 + \frac{4}{10}\cdot 3 + \frac{1}{10} \cdot 2 \\
	& = 4 \cdot \frac{5}{10} + 3 \cdot \frac{4}{10} + 2 \cdot \frac{1}{10} \\
	& =  3.4 \;.
\end{align*}
</div>

## Expected Value

### The big idea

The key idea here is that the mean is a ________ of ___________ times _______________.
<!-- \textbf{sum of values times probabilities}.   -->

\bigskip

We can write this as
\[
\mbox{mean} = \sum \mbox{value} \cdot \mbox{probability}
\]

### Discrete Random Variables

For a discrete random variable this translates to
\[
\mu_X = \E(X) = \sum x f(x)
\]
where the sum is taken over all possible values of $X$ and $f$ is the pmf for $X$.

The mean of a random variable also goes by another name: \textbf{expected value}.  We 
can denote this as $\mu_X$ or $\E(X)$.

**Example 2.**
Let $X$ be discrete random variable with probablilities given in the table 
below.


------------------------------------
   value of $X$      0     1     2    
------------------ ----- ----- ----- 
 **probability**    0.2   0.5   0.3
------------------------------------

What is the mean (expected value) of $X$?

<div class = "explain" label = "Show solution">
$E(X) = 0 \cdot 0.2 + 1 \cdot 0.5 + 2 \cdot 0.3 = 0.5 + 0.6 = 1.1$
This value reflects the fact that the random variable is larger than
1 a bit more often than it is less than 1.
</div>
<br>
<br>

\vfill
\vfill

\newpage

**Example 3.**
A local charity is holding a raffle.  They are selling $1000$ raffle tickets 
for $5 each.  The owners of five of the raffle tickets will win a prize.  The 
five prizes are valued at $25, $50, $100, $1000, and $2000.  
Let $X$ be the value of the prize associated with a random raffle ticket 
($0 for non-winning tickets). 

a. What is the probability of winning a prize?

b. What is the probability of winning the grand prize?

c. What is the expected value of the prize?

<div class = "explain" label = "Show solution">

* $\Prob{\mbox{the ticket wins a prize}} = \Prob(X > 0) = 5/1000$.
*	$\Prob{\mbox{the ticket wins the grand prize}} = \Prob(X = 2000) = 1/1000$.
*	$\Prob{\mbox{the ticket wins a prize worth more than} 75 dollars} = \Prob(X > 75) = 3/1000$.

The expected value of a ticket is

\[
  0 \cdot \frac{995}{1000}
  + 25 \cdot \frac{1}{1000}
  + 50 \cdot \frac{1}{1000}
  + 100 \cdot \frac{1}{1000}
  + 1000 \cdot \frac{1}{1000}
  + 2000 \cdot \frac{1}{1000}
\]

```{r }
25 * .001 + 50 * 0.001 + 100 * 0.001 + 1000 * 0.001 + 2000 * 0.001
# R can help us set up this sum:
sum(c(25, 50, 100, 1000, 2000) * 0.001)
```

</div>
<br>
<br>

\vfill
\vfill
\vfill

### Continuous Random Variables

When working with a continuous random variable, 
we replace the sum with an integral and replace the 
probabilities with our density function to get the following definition:

\[
\E(X) = \mu_X = \int_{-\infty}^{\infty} x f(x) \; dx
\]

**Example 4.**
Calculate	the mean of a symmetric triangle distribution on $[-1, 1]$.

```{r echo = FALSE, fig.width = 5, fig.height = 1.5}
library(triangle)
gf_function(dtriangle, args = list(a = -1, b = 1, c = 0), xlim = c(-2, 2))
```

<div class = "explain" label = "Show solution">	
\begin{align*}
	\E(X) & = \int_{-1}^{1} x f(x) \; dx 
	\\
	& = \int_{-1}^{0} x (x - 1) \; dx + \int_{0}^1 x (1 - x) \; dx 
	\\
	& = \int_{-1}^{0} (x^2 - x) \; dx + \int_{0}^1 (x - x^2) \; dx  
	\\
	& = \left.\frac{x^3}{3} - \frac{x^2}{2} \right|_{-1}^0
	+ \left. \frac{x^2}{2} - \frac{x^3}{3} \right|_{0}^1
	\\
	& = \frac13 - \frac12 + \frac12 - \frac13 = 0
	\\
\end{align*}
<br>
<br>

\newpage

We could also calculate this numerically in R:

```{r fig.width = 5, fig.height = 1.5}
library(mosaicCalc)
# define the function -- (abs(x) <= 1) will converted to 0 or 1
f <- makeFun( (1 - abs(x)) * (abs(x) <= 1) ~ x)

# plot the function to make sure it looks right
gf_function(f, xlim = c(-2, 2))   

# define x * f(x) as another function
xf <- makeFun( x * f(x) ~ x )
# integrate over the interval from -1 to 1 to get E(X)
integrate(xf, -1, 1)

# use anti-derivative instead -- lower.bound sets the "zero point" for the anti-derivative
F <- antiD( x * f(x) ~ x, lower.bound = -1)
F(-1)  # should be 0 because of lower.bound
F(1)   # should be the expected value -- also 0.
```
</div>
<br>
<br>

\newpage

## Variance

Arguing similarly, we can compute the variance of a discrete or continuous random
variable using 

* discrete:
$\Var(X) = \sigma^2_X = \sum_x (x-\mu_X)^2 f(x) \; dx$

* continuous:
$\Var(X) = \sigma^2_X = \int_{-\infty}^{\infty} (x-\mu_X)^2 f(x) \; dx$

These can be combined into a single definition by writing
$$
\Var(X) = \E((X - \mu_X)^2) \;.
$$
Note: It is possible that the sum or integral used to define the mean (or the variance) will fail 
to converge. In that case, we say that the random variable has no mean (or variance) or that the 
mean (or variance)  fails to exist.[^converge]

[^converge]: Actually, we will require that
$\int_{\infty}^{\infty} |x| f(x) \; dx$ converges and
$\int_{\infty}^{\infty} |x|^2 f(x) \; dx$ converges.
If these integrals (or the corresponding sums for discrete random variables) fail to converge, 
we will say that the distribution has no mean (or variance).

**Example 5.**
Compute the variance of symmetric triangle distribution on $[-1,1]$.

\vfill
\vfill

<div class = "explain" label = "Show solution">	
```{r}	
f <- makeFun( (1 - abs(x)) * (abs(x) <= 1) ~ x)
xxf <- makeFun( (x-0)^2 * f(x) ~ x )
integrate(xxf, -1, 1)
G <- antiD( (x-0)^2 * f(x) ~ x)
G(1) - G(-1)
```
</div>
<br>
<br>


\newpage

### Useful identity

Some simple algebraic manipulations of the sum or integral above shows that

\begin{align}
\Var(X) &= \E(X^2) - \E(X)^2
\end{align}


**Example 6.**
Compute the mean and variance of the random variable with pdf given by
	
$$
g(x) = \frac{3x^2}{8} [\![ \; x \in [0,2] \;  ]\!] \;. 
$$

\vfill
\vfill

<div class = "explain" label = "Show solution">	
```{r}	
g <- makeFun( (3 * x^2/8 ) * (0 <= x & x <= 2) ~ x )
m <- antiD( x * g(x) ~ x, lower.bound = 0)(2)  # all in one step instead of defining F or G
m
v <- antiD( (x - m)^2 * g(x) ~ x, m = m, lower.bound = 0)(2)
v
# here's the alternate computation
antiD( x^2 * g(x) ~ x, lower.bound = 0)(2) - m^2
```
</div>
<br>
<br>

\bigskip

As with data, the standard deviation is the square root of the variance.



\newpage

## More Practice

**1.** You are invited to play the following game. 
You draw two chips without replacement from a jar containing 5 red, 5 blue, and
5 green chips.   If both chips have the same color you win $5.  If the two chips
have different colors, you win $3.  On average, how much will you win per game?
It costs you $4 to play the game.  On average, would you win money, lose money,
or break even playing this game?

<!-- x    	$5		$3 -->
<!-- p(x)  	.286		.714 -->

<!-- On average you will win 5*.286 + 3*.714 = $3.57.  On average you will lose playing this game. -->

**2.** Toss a fair coin 3 times. 
Let X be the number of heads produced.  Find the pmf for $X$ and
use it to find the average number of heads produced when the coin is tossed 3
times.

<!-- x		0	1	2	3 -->
<!-- p(x)	1/8	3/8	3/8	1/8 -->

<!-- E(X) = (0+3+6+3)/8 = 3/2 = 1.5. -->

**3.**	Repeat (3) above under the assumption that the coin is biased and only has a 1/4 
chance of producing a head.

<!-- x		0		1		2		  3 -->
<!-- p(x)	27/64		27/64		9/64		1/64 -->

<!-- E(X) = (27+18+3)64 = 48/64 = ¾ = .75. -->

**4.**	Let 
$f(x) = 2e^{-2x}$ on $(0, \infty)$.
Create this function in R and use R to integrate it on $[0, \infty)$.
Note that $e^x$ is expressed in R as `exp(x)`. 

a. Is $f(x)$ is a pdf?

```{r include = FALSE}
f <- function(x) 2 * exp(-2*x) * (x>=0)
integrate(f,0,Inf)
```

b. Use R to compute $\Prob(-1 \le X \le 3)$.

```{r include = FALSE}
integrate(f,-1,3)
```

c. Use R to compute the cumulative distribution function $F$ and use $F$ to compute 
$\Prob(3 \le X \le 10)$.

```{r include = FALSE}
F <- antiD(f(x) ~ x, lower.bound = 0)
F(10) - F(3)
```

d. Use R to compute the mean of $X$.

```{r include = FALSE}
xf <- function(x) x * f(x)
integrate(xf,0,Inf)
```

**5.**
Let $X$ have the pdf $f(x) = 2x$ on $[0, 1]$.  

a. Find the median of $X$. [Hint: What is a median? -- Don't think about sorting.]

b. What is the 20th percentile of $X$?


**6.**
Do the following by hand, without using R.
Let $f(x) = C x^3$ on $[0, 3]$ be the pdf for $X$.

a. What is the value of $C$?

b. Compute $\E(X)$.

c. Compute $\Prob(X > 2)$.

d. Determine the median of $X$.





