---
title: "Testing Statistical Hypotheses"
author: "Stat 241"
output:
  html_document:
    theme: yeti
    css: ['../styles/ds303-notes.css', '../styles/worksheets.css']
  df_document: default
  pdf_document: default
---

\def\tor{\operatorname{or}}
\def\tand{\operatorname{and}}
\def\tnot{\operatorname{not}}
\def\Prob{\operatorname{P}}
\def\E{\operatorname{E}}
\def\Var{\operatorname{Var}}
\def\SD{\operatorname{SD}}
\def\Binom{{\sf Binom}}
\def\Norm{{\sf Norm}}
\def\T{{\sf T}}
\def\Chisq{{\sf Chisq}}
\def\Unif{{\sf Unif}}
\def\Exp{{\sf Exp}}
\def\Beta{{\sf Beta}}
\def\Gamm{{\sf Gamma}}
\def\Weibull{{\sf Weibull}}
\def\Pois{{\sf Pois}}

<script src="../styles/folding.js"></script>

```{r hooks, message=FALSE, include = FALSE}
knitr::knit_hooks$set(fold = function(before, options, envir) {
  if (before) {
    return(glue::glue('<div fold = "{options$fold}">'))
  } else {
    return('</div>\n')
  }
})
```


```{r setup, include=FALSE}
onLine <- TRUE
showSols <- TRUE
knitr::opts_chunk$set(
  results = "hold",
  echo = TRUE,
  fig.width = 5, 
  fig.height = 2,
  fig.align = "center"
)
library(mosaic)
library(pander)
library(mosaicCalc)
library(triangle)
library(patchwork)
library(reactable)
theme_set(theme_bw(base_size = 8))
set.seed(20210108)
options(digits = 3)
```
 

## What's on the FE Exam?

Depends a bit on your specialty.  Here's what it says for Mechanical Engineering:

* Numerical methods (e.g., approximations, precision limits, 
**error propagation**, Taylor's series, Newton's method)

* Probability distributions (e.g., normal, binomial, empirical, discrete, continuous)

*  Measures of central tendencies and dispersions (e.g., mean, mode, standard deviation, 
confidence intervals)

* Expected value (weighted average) in decision making

* Regression (linear, multiple), curve fitting, and goodness of fit
(e.g., correlation coefficient, least squares)

## Some Sample Problems

Sources:

* G.K. Michaelson, Marshall Univ., youtube
* Fundamentals of Engineering, D.G. Newnan (FE Exam Preparation)
* F. Haan (ME321, CE356)
* Notes, solutions, comments, and some adjustements by R Pruim

<br>

<div class = "example">
**1.**
What is most nearly the sample
standard deviation of the following data set? {4, 6, 9, 9}

&nbsp;
   a) 2.1 &nbsp;   b) 2.4 &nbsp;   c) 4.5 &nbsp;  d) 6.0 &nbsp;

<br>

**Notes**

* What can you rule out without doing an detailed computations?
* What error are they checking for with the remaining options?
* What error are they checking for with the options you ruled out?

<div class = "explain" label = "Show solution">
```{r}
x <- c(4, 6, 9, 9)
sd(x)
var(x) 
sum( (x - mean(x))^2 / 3)
sum( (x - mean(x))^2 / 4)
sum( (x - mean(x))^2 / 4) %>% sqrt()
```
</div>
</div>

<br>
<div class = "example">
**2.**
Four fair dice are thrown.  What is the probability
of rolling at least one six?

<div class = "explain" label = "Show solution">
```{r include = showSols}
# 1 minus no sixes
1 - (5/6)^4
1 - dbinom(0, 4, 1/6)
```
</div>

</div>

<br>
<div class = "example">
**3.**
A bag contains 100 chips numbered 1 to 100.
One chip is drawn from the bag.
What is the probability that the number on the chip selected
will be odd or greater than 80?

&nbsp;
   a) 0.1 &nbsp;
   b) 0.5 &nbsp;
   c) 0.6 &nbsp;
   d) 0.7 &nbsp;

<br>

**Notes**

* What error or problem modification could lead to each of the wrong
answers?

<div class = "explain" label = "Show solution">

Key thing: $\Prob(A \tor B) = \Prob(A) + \Prob(B) - \Prob(A \tand B)$

P(odd) = 0.5; 

P(greater than 80) = 0.2; 

0.5 + 0.2 = 0.7

P(both) = 0.1; 

P(one or the other) = 0.5 + 0.2 - 0.1 = 0.6

You could also do this by counting how many of the 100 numbers are either odd or
greater than 80.

</div>
</div>

<br>
<div class = "example">
**4.**
The final scores on a test are normally distributed
with a mean of 72 and a standard deviation of 10.
What is most nearly the probability that a
student's score will be between 65 and 78?

&nbsp;
   a) 0.4196 &nbsp; 
   b) 0.4837 &nbsp; 
   c) 0.5161 &nbsp; 
   d) 0.6455 &nbsp; 

<br>

**Notes**

* For the FE exam you will have access to a normal probability table
like [this one](https://d2vlcm61l7u1fs.cloudfront.net/media%2F4aa%2F4aa0d140-7822-422e-a489-54983576873f%2FphpLwzeqs.png). Do this problem using that 
table. (Then check your work in R.)

* Each column of the table could be computed in R, how would you do it?
(Below is a fancy version that formats everything into a pageable table. Don't worry
about that part, just see if you know how to calculate any particular value in the 
table using R.)

```{r, results = "hold", fold = "so", R.options = list(digits = 4)}
tibble(
  x = seq(0, 3, by = 0.1),  # make a sequence from 0 to 3 by tenths
  `f(x)` = dnorm(x),
  `F(x)` = pnorm(x),
  `R(x)` =  1 - pnorm(x),
  `W(x)` = pnorm(x) - pnorm(-x)
) %>% 
  round(4) %>%
  reactable::reactable()   # fancy table
```

<div class = "explain" label = "Show solution">
```{r}
pnorm(78, mean = 72, sd = 10) - pnorm(65, mean = 72, sd = 10) 
# if we assume integer scores that are approximatley normal, 
# we might use a "continuity correcttion"
pnorm(78.5, mean = 72, sd = 10) - pnorm(64.5, mean = 72, sd = 10) 
```

</div>
</div>

<br>
<div class = "example">
**5.**
Four fair coins are tossed once.  What is the probability of 
obtaining three heads and one tails?

&nbsp;
  a) 1/4 &nbsp;
  b) 3/8 &nbsp;
  c) 1/2 &nbsp;
  d) 3/4 &nbsp;
 

<div class = "explain" label = "Show solution">

There are only 16 possible outcomes, so you could list them all and count.
But using what we know of binomial distributions is less tedious (and scales
to problems of larger size.

```{r fold = 'so'}
dbinom(1, 4, 0.5)
4 * 0.5^1 * 0.5^3
choose(4, 1) * 0.5^1 * 0.5^3
```

</div>
</div>

<br>
<div class = "example">
**6.**
How many teams of 4 can be formed from 35 people?

&nbsp;
  a) about 25,000 &nbsp;
  b) about 2,000,000 &nbsp;
  c) about 50,000 &nbsp;
  d) about 200,000 &nbsp;


<br>

**Notes**

* We haven't done problems exactly like this, 
but we did learn the key thing you need: 
binomial  coefficients. With that hint, can you solve this?


<div class = "explain" label = "Show solution">
```{r}
choose(35, 4)
factorial(35) / (factorial(4) * factorial(35 -4))
```
</div>
</div>

<div class = "example">
**7.**  
How many three-letter codes can be formed from the English alphabet if no repetitions are allowed?

&nbsp;
  a) $26^3$ &nbsp;
  b) $26/3$ &nbsp;
  c) $(26)(25)(24)$ &nbsp;
  d) $26^3/3$ &nbsp;

<br>

**Notes**

* This is related to the following probability problem: If a random 
three-letter code is selected (not repetitions allowed), what is 
the probability of selecting ABC?

* The State of Michigan runs a lottery game that is similar to this, but uses
digits rather than letters and allows for repetition. (It's called Pick 3.)
If you select 123, what is the probability that your lottery ticket is the 
winner?

<div class = "explain" label = "Show solution">
```{r}
# probability problem
prob <- 1/26 * 1/25 * 1/24; prob
# 1/prob is the number of possible codes 
1/prob
# alternative solution
26 * 25  * 24
```
</div>
</div>

<br>
<div class = "example">
**8.**
There are 10 defectives per 1000 products. What is the probability that there is
one and only one defective in a random lot of 100?

&nbsp;
  a) $99 \cdot 0.01^{99}$ &nbsp;
  b) $0.01$ &nbsp;
  c) $0.5$ &nbsp;
  d) $0.99^{99}$ &nbsp;

<br>

**Notes**

* Given the options here, one is **_approximately_** but not **_exactly_** correct.
  Why isn't it exactly correct? Why is it a pretty good approximation.
  
* What would the the exact probability be for this? (It's a bit of a 
pain to compute unless you are clever in your use of R, so you don't have
to actually compute it.)

<div class = "explain" label = "Show solution">
```{r results = 'hold', fold = 'so'}
# Binomial approximation
100 * 0.01 * .99^99
dbinom(1, 100, 0.01)
```

```{r results = 'hold', fold = 'so'}
# Exact method
100 * 10/1000 * prod((990:892)/(999:901))
# this situation is is modeled by the Hypergeometric distribution
# R has a function for that distribution as well
dhyper(1, 10, 990, 100)
```
</div>

</div>

<br>
<div class = "example">
**9.**
A set of six power measurements was made. The mean $\dot{𝑄}$̇ was 22450 W. 
The standard deviation from the six measurements was 846 W. 
Estimate the 95% confidence interval about the mean.

&nbsp;
  a) $22450 \pm 691$ W &nbsp;
  b) $22450 \pm 888$ W &nbsp;
  c) $22450 \pm 1036$ W &nbsp;
  d) $22450 \pm 1210$ W &nbsp;


<div class = "explain" label = "Show solution">

**Note:** The FE exam has a t-distribution table too. It is a bit different.
It is designed for computing critical values primarily.  So you can look up
the degrees of freedom and desired tail probability to get the value of $t_*$.
[Here's the table](FE-reference.pdf#page=49)

```{r}
# correct solution with t distribution
SE <- 846 / sqrt(6)
t.star <- qt(0.975, df = 5); t.star
SE * t.star

# if you incorrectly use the normal distribution (this explain choice a)
SE * 1.96
SE * 2
```

</div>
</div>

<br>
<div class = "example">
**10.**
An auto manufacturer has three plants (A, B, and C). Four out of 500 cars from
Plant A must be recalled, 10 out of 800 from Plant B and 10 out of 1000 from
Plant C. Now a customer purchases a car from a dealer who gets 30% of his stock
from Plant A, 40% from Plant B, and 30% from Plant C, and the car is recalled.
What is the probability it was manufactured in Plant A?

&nbsp;
  a) 0.0008 &nbsp;
  b) 0.010 &nbsp;
  c) 0.0125 &nbsp;
  d) 0.231 &nbsp;
  
<div class = "explain" label = "Show solution">
```{r}
0.3 * 4/500 / (
  0.3 * 4/500 + 
  0.4 * 10/800 +
  0.3 * 10/1000
)
```
</div>

</div>

<br>
<div class = "example">
**11.**
Linear rgression is ued to plot a straight line through the following
data points.

```{r echo = FALSE}
DDD <-
  tibble(
  x=c(1, 2, 3, 5),
  y = c(6, 7, 11, 13)
) 
DDD %>% pander()
```

What is most nearly the slope of the regression line?

<br>

**Notes**

* I hope you don't get a problem like this one. But given enough time, you could compute the answer.

  * What fact about the regression line could you use?
  * Do you know how to do all the computations involved? (There is one 
  part of this we never did by hand, but on such a small data set it is
  possible, although tedious.)
  * Can you estimate any of the components of this calculation well
  enough to eliminate any of the options.
  
* Make a (hand-drawn) sketch. How many answers can you rule out just from
your sketch?

* For the actual FE test, you get a [reference book](FE-reference.pdf#page36).

  * Take a look at it. How much do you recognize?
  
  * Notice that all the "nasty formulas" for the standard errors that R computes 
  for you are there.  Those are really tedious to use by hand.
  
  * The formulas presented don't really promote good statistical thinking. 
  Several of them are chosen to reduce the amount of button pushing on a calculator
  and take advantage of some algebraic identities. Compare the formula for 
  the correlation coefficient given in these materials vs the one we saw in class.
  Which one is easier to remember? Which one tells you more about what the 
  correlation coefficient is doing?
  
      * Note: There is a really nice connection between $R$ and the angle between two
  vectors; so there is some really nice linear algebra involved.  Take Stat 344
  or Math 355 if you want to know more about this.
  

<div class = "explain" label = "Show solution">
```{r echo = FALSE}
lm( y ~ x, data = DDD) %>% broom::tidy() %>% pander()
```
</div>
