---
title: "2. Summarizing Data with Numbers"
author: "Stat 241"
output:
  pdf_document: default
  html_document:
    theme: yeti
---


```{r, setup, include = FALSE}
spaces <- function(n) {
  paste(rep(" ", n), collapse = "")
}
library(mosaic)
library(palmerpenguins)
theme_set(theme_bw(base_size = 7))
knitr::opts_chunk$set(
  fig.align = "center",
  fig.width = 4, 
  fig.height = 2
)
library(patchwork)

showSols <- FALSE
```

## Summarizing quantitative data

Generally we are interested in measures of **center** (what is a "typical" value?) and 
**spread** (are the values clustered close together or more spread out?).

###	Five Number Summary and Boxplots

For a set of numerical data:

*	Median (Q2):

\vfill

*	First Quartile (Q1):

\vfill

*	Third Quartile (Q3):

\vfill

*	Min (Q0):

\vfill

*	Max (Q4):

\vfill

*	Interquartile Range (IQR):

\vfill

* Five Number Summary:  
<!-- min, Q1, median, Q3, max (ie, Q0, Q1, Q2, Q3, Q4) -->

\vfill

### By Hand Example 

1. Compute the numerical and graphical summaries listed for the following data values:  
$3, 5, 2, 3, 7, 5, 8, 6, 7$

    a. Five Number Summary: \vfill \vfill

    b. IQR:   \vfill

    c. Boxplot: \vfill

\newpage

### Five Number Summary Using R  

**Example:** Find the Five Number Summary for the heights of the adult children in the data frame Galton.

```{r, fig.width = 4, fig.height = 1.0}
quantile( ~ height, data = Galton)
gf_boxplot( ~ height, data = Galton)
```

### Boxplots and skew

2. Sketch a boxplot for each of the following situatoins.

    a.  Boxplot of a data set skewed to right. \vfill
    b.  Boxplot of a data set skewed to left. \vfill

## Mean and Standard Deviation

### Some notation 

* $n$ = number of data values
*	$x$ = variable
*	$x_1, x_2, \dots x_n$ = individual values of the variable $x$.

###	Mean
		 
* mean of $x$ = $\displaystyle \overline{x} = \frac{\sum_{i=1}^n x_i}{n}$

  * "add them all up and divide by how many there are"

#### Example Computation

3. Compute the mean for $x: 1,1,3,4,6,7,7,9$.

\vfill
\vfill

\newpage

#### Things to note about the mean and median

* If the distribution is symmetric...  \vfill

* If the distribution is skewed to the right...  \vfill

* If the distribution is skewed to the left...  \vfill

* If there are outliers in the data...  \vfill

* If the distribution is not unimodal...  \vfill

## Variance and Standard Deviation

* sample variance of x = $\displaystyle s_x^2 = \frac{ \sum_{i=1}^{n} (x_i - \overline{x})^2 }{n-1}$
    * "add up squared differences from the mean and divide by 1 less than how many there are"
    * "sample" indicates that we only have data for a sample, not for all individuals; since that
    is the usual situation, we will drop the word sample.
* sample standard deviation of x = $\displaystyle s_x = \sqrt{\sum_{i=1}^{n} \frac{ (x_i - \overline{x})^2}{n-1}}$

* Why do we square the differences?

  * If we do nothing...  \vfill
  
  * If we use absolute value... \vfill

\newpage

### Small Example:  

4. Compute the variance and standard deviation for $x$: $1,2,3,4,5$

\vfill
\vfill

## Computing summary statistics with R

We won't usually compute these statistics by hand.  The hand computations above are just
to make sure we understand what these numbers are. But R can quickly and easily compute
any of these statistics (and more) with the following template.

```{r eval = FALSE}
stat( ~ x, data = mydata)
```

We can compute statistics within groups using either of these forms:

```{r eval = FALSE}
stat( ~ x | z, data = mydata)
stat( y ~ x, data = mydata)
```

### Example: Penguin body mass

```{r}
library(palmerpenguins)
df_stats( ~ body_mass_g, data = penguins)
```

Notice that there are some missing values (penguins for which no body mass was recorded).
This causes problems for some of our statistics. We can't really compute the mean of all of the 
penguins if we don't know all of the values.  `df_stats()` lets us know that there are some missing
values and that our statistics are computed only on the non-missing values. But the individual
statistic functions report `NA` (not available, missing).

```{r}
mean( ~ body_mass_g, data = penguins)
```

\newpage

We can fix this two ways:

```{r}
mean( ~ body_mass_g, data = penguins, na.rm = TRUE)
df_stats( ~ body_mass_g, data = penguins, mean)
```

Let's end this example by comparing the body mass of the different penguin species.
Notice the different output formats below.

```{r}
mean( ~ body_mass_g | species, data = penguins, na.rm = TRUE)
df_stats( ~ body_mass_g | species, data = penguins, mean)
```

###  Example: HELPrct

The data frame HELPrct contains data on an experiment testing a treatment of
substance abusers. It is too large to conveniently display the entire data set, but 
we dan see the top few rows of the data using `head()`.

```{r}
head(HELPrct)
```

That's a bit ugly because there are so many variables.  Let's try `glimpse()` instead


```{r}
glimpse(HELPrct)
```


`inspect()` computes some summary statistics for each variable instead of
showing the raw data (for a few rows).  Give it a try:

```{r, eval = FALSE}
inspect(HELPrct)
```

We can get summary statistics about age for each of the three abuse groups.

```{r}
df_stats(age ~ substance, data = HELPrct)
df_stats( ~ age | substance, data = HELPrct)
```

## Histograms and boxplots by groups

```{r fig.width = 5, fig.height = 4}
gf_histogram( ~ age | substance ~ ., data = HELPrct)
```

```{r fig.width = 3, fig.height = 2.5}
gf_boxplot( age ~ substance, data = HELPrct)
```
  
## Quantitative vs Categorical Variables

A quantitative variable is a variable that takes on numerical values in such a
way that it makes sense to perform arithmetic operations on its values.  A
categorical variable is a variable whose values are categories. 
Body mass (in grams) is a quantitative variable;  
`species` is a categorical variable.  

Sometimes numbers are used to represent categories.
Using numbers to represent categories does not turn a
categorical variable into a quantitative variable, but it can confuse R,
which will treat the variable as quantitative if we use numbers.
We can explicitly tell R we have a categorical variable using `factor(x)`.
(Factor is R's lingo for a categorial variable, it has nothing to do with
factoring numbers.)

### Example: Kids feet

The data frame `KidsFeet` contains data on thirty-nine 4th grade children

```{r}
head(KidsFeet)
```

5. Which variables are quantitative and which are categorical?

2. What sorts of plots are appropriate to look at the distribution of a quantitative variable?
Make a few of these for some of the quantitative variables.

3. What sorts of plots are appropriate to look at the distribution of a categorical variable?
Make a few of these for some of the categorical variables.

4. Are birth month and birth year categorical or quantitative?

5. These plots probably surprise you. 
Can you figure out what R is doing? How can you fix them?

```{r, indent = spaces(4), fig.width = 4, fig.height = 2, warning = FALSE}
gf_boxplot(length ~ birthyear, data = KidsFeet) |
gf_boxplot(birthyear ~ length, data = KidsFeet)
```

6. Compute the mean foot length separately for boys and girls.

7. Create several different plots showing the distribution of foot lengths
separately for boys and girls. Whic one do you prefer? Why?
 
## EXERCISES 

**2.1**	In the data frame `HELPrct`, the variable `avg_drinks` (also called `i1`) gives the average number 
of drinks per day, `substance` gives the kind of substance abuse, and `sex` gives the gender of the
subject.

a. Create histograms comparing the average number of drinks per day by gender.
b. Repeat (a) with boxplots.
c. Compute the means and standard deviations for the daily number of drinks for male and for females.  
d. Based on the information in (c), would you conclude that, on average, the males in the study 
consume more alcohol than the females?  
e. For which gender is there more variability in alcohol consumption?
f. Create frequency histograms comparing the alcohol consumption by type of substance abuse.  
g. Is the distribution of alcohol consumption symmetric for those whose are alcohol abusers?  
   If it is skewed, in which direction is it skewed? 
h. Compute the median alcohol consumption for each of the three types of abuse.

```{r include = showSols, R.options=(list(digits = 2))}
gf_dhistogram( ~ avg_drinks | sex ~ ., data = HELPrct)
gf_histogram( ~ avg_drinks | sex ~ ., data = HELPrct)
gf_boxplot( avg_drinks ~ sex, data = HELPrct)
df_stats( avg_drinks ~ sex, data = HELPrct, mean, sd)
gf_histogram( ~ avg_drinks | substance ~ ., data = HELPrct)
df_stats( avg_drinks ~ substance, data = HELPrct, median)
```

**2.2**	Compute the mean, variance and standard deviation for the data set $x: 1,5,3,7,9$  by HAND.

```{r include = showSols}
# these should be done by hand, but here are the answers
x <- c(1,5,3,7,9)
mean(x)
var(x)
sd(x)
```

**2.3**	Create a set of 6 numbers in the range 0 -– 10 (inclusive) that will have the largest possible 
standard deviation. 

```{r include = showSols}
# largest possible
df_stats( ~ c(0, 0, 0, 10, 10, 10), data = data.frame(), var, sd, mean)
# not as large, but uses unique values (not required, but some students might assume this)
df_stats( ~ c(0, 1, 2, 8, 9, 10), data = data.frame(), var, sd, mean)
```

**2.4**	Below is the boxplot of a data set.

```{r indent = spaces(0), echo - FALSE, fig.height = 1, fig.width = 5}
set.seed(1234)
DDD <- tibble(x = rgamma(50, shape = 1.5, scale = 10))
gf_boxplot( "" ~ x, data = DDD) %>%
  gf_labs(y = "", x = "")
```

a. What is the median of the distribution?
b. What is the IQR for the distribution?
c. Is the mean larger or smaller than the median?  How do you know?


```{r include = showSols, R.options = list(digits = 2)}
df_stats( ~ x, data = DDD, median, iqr, mean) 
```

**2.5**	From the data frame HELPrct, create a bargraph that shows how 
homelessness (`homeless`) is distributed with respect to (`sex`). 
Produce the version where the bars are adjacent (dodged).
Should you use counts or proportions? Why? If proportions, what 
should the denominator be?

Based on the picture, should you conclude that males and females in the study are 
homeless at roughly the same rate or at different rates?

```{r include = showSols}
# we want bars to sum to 1 for male and for female
gf_props( ~ sex, data = HELPrct, fill = ~ homeless, position = "dodge", denom = ~ x)
```