---
title: "3. Simulating for Probability"
author: "Stat 241"
output:
  html_document:
    theme: yeti
    css: ../styles/ds303-notes.css
  df_document: default
  
---


<script src="../styles/folding.js"></script>

```{r hooks, message=FALSE, include = FALSE}
knitr::knit_hooks$set(fold = function(before, options, envir) {
  if (before) {
    return(glue::glue('<div fold = "{options$fold}">'))
  } else {
    return('</div>\n')
  }
})
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 5, 
  fig.height = 2,
  fig.align = "center"
)
library(mosaic)
library(pander)
theme_set(theme_bw())
set.seed(20210108)
```


Goal: Learn how to use R to simulate flipping coins, rolling dice, and dealing cards -- three common
examples used in probability examples.

## Flipping coins

To simulate flipping coins, we will use two important R functions:

* `rflip(n)` simulates simulates `n` tosses of a fair coin.

* `do(n) * soemthing` repeats something `n` times. (Read it as "do n times"...)

<br />

### Example 1 

Toss a fair coin 20 times.  Let $X$ be the number of heads.

```{r}
rflip(20)
```

Let's do that several times

```{r}
do(5) * rflip(20)
```

Notice that `do()` records the information in a nice summarized format.

If we want to estimate probability, we need to repeat this many times. Let's do it 10,000 times.

```{r}
Flips20 <- do(10000) * rflip(20)
```

Let's look at the first few rows to make sure we know what we have.[^pander]

[^pander]: The `pander()` function is in the `pander` package. It creates prettier output for 
things like tables and data frames.

```{r}
head(Flips20) %>% pander()
```

We can let R count the number of times each value from 0 to 20 is produced by our 10000 simulations
and present the information to us in a table or as a histogram

```{r} 
tally( ~ heads, data = Flips20) %>% pander()
gf_histogram( ~ heads, data = Flips20, binwidth = 1)
```

Perhaps we would rather have proportions rather than counts.

```{r}
tally( ~ heads, data = Flips20, format = "prop") %>% pander()
gf_histogram( ~ heads, data = Flips20, binwidth = 1)
```

The table above allows to to calculate probabilities about the number of heads in 20 tosses of a fair coin.

<br />

### Example 2

Toss a biased coin 10 times and let $X$ = number of heads.  
Assume the biased coin has a probability of 0.25 of producing a head.

We simulate 10 tosses of the biased coin.

```{r}
rflip(10, prob = 0.25)
```

Now let's repeat that 10,000 times, find the estimated probability distribution for $X$, and make a histogram.

```{r}
Flips10 <- do(10000) * rflip(10, prob=.25)
tally( ~ heads, data = Flips10, format ="prop") %>% pander()
gf_dhistogram( ~ heads, data = Flips10, binwidth = 1)
```

**Q.** What is (approximately) the probability of getting at least 5 heads?

We could answer this by doing the arithmetic ourselves from the table above, but we should let 
R do that for us.  Here's how.

```{r}
tally( ~(heads >= 5), data = Flips10)
```

#### Your Turn

What is the probability of getting 3 or fewer heads?

\vfill

```{r, include = FALSE}
tally( ~(heads <= 3), data = Flips10)
```


<br />

### Example 3	

We can use coin tosses to simulate anything that has two possible outcomes. 
Here's an example about shooting free throws.

Jo has a free throw shooting average of 85%.  She has been fouled in the act of
shooting a 3-point shot as time runs out. Jo has 3 free throws coming. Her
team is behind by 2 points.  Estimate the probability that her team does not
lose in regulation time. Use 10000 simulations.

```{r, fold = "so"}
FreeThrows <- do(10000) * rflip(3, prob = 0.85)
tally( ~(heads >= 2), data = FreeThrows, format = "prop") 
```


## Rolling dice

### Example 4  

Roll of a pair of fair dice. Let $S$ = sum of the results.

```{r}
1:6
```

We can generate the result of tossing a fair die 10 times

```{r}
resample(1:6, 10)
```

We now simulate 10000 tosses of a fair die twice.[^tibble]

[^tibble]: The `tibble()` function creates data frames. You simply specify the values
for each of the variables in the data frame. In this case, we are creating `die1` and `die2` 
randomly, and then creating `S` by adding those two values together.

```{r}
TwoDice <-
  tibble(
    die1 = resample(1:6, 10000),
    die2 = resample(1:6, 10000),
    Sum = die1 + die2)                # add them together, componentwise
head(TwoDice) %>% pander()
```

We create the estimated probability distribution for $S$ and the histogram.

```{r}
tally( ~ Sum, data = TwoDice, format = "prop") %>% pander()
gf_dhistogram( ~ Sum, data = TwoDice, binwidth = 1)
```


### Your turn

**1.** A pair of fair dice are tossed. 
Use 10000 simulations to estimate the probability that the dice turn up with the same value.

<div class="explain" label = "Show solution">
```{r}
tally( ~(die1 == die2), data = TwoDice, format = "prop") %>% pander()
```
</div>


**2.** Estimate the probability that the sum of two dice is more than 9.

<div class="explain" label = "Show solution">
```{r}
tally( ~(Sum > 9), data = TwoDice, format = "prop")
```
</div>


## Dealing cards

An ordinary deck of playing cards contains 52 cards with 4 suits (clubs,
diamonds, hearts, spades) of 13 cards each (2 -- 10, Jack, Queen, King, and Ace).
The clubs and spades are black suits; hearts and diamonds are red.

We can simulate a deck of cards using the numbers 1:52. (The `mosaic` package
includes a slightly fancier thing called `Cards`, but it is harder to do 
arithmetic with that.)

For example, here are two 5-card hands.

```{r}
set.seed(1234)          # so we get the same random draw each time we run this
sample(1:52, 5)
sample(Cards, 5)        # 9S = 9 of spades; etc.
```

Note that we are using `sample()` here rather than `resample()`.  
`resample()` allows for  a value to be repeated ("sampling with replacement). 
But with cards, you can only get each card once.  
`sample()` does not allow for repetition ("samping without replacement").
This ensures that we get 5 different cards each time.

<br />

### Example 5

If we deal 3 cards from a standard deck, what is the probability that all three are spades? 

Let's represent spades as the cards 1--13. So the questions becomes what is the probability
of getting three numbers in that range. R will create TRUE and FALSE values if we do a comparison,
and `sum()` will treat TRUE like 1 and FALSE like 0.  Putting that all together, it is pretty
easy to count how many spades are in a hand of three cards.

```{r fold = "o"}
hand1 <- sample(1:52, 3) 
hand1
hand1 <= 13
sum(hand1 <= 13)
```

We can put this altogether in one line.  That will make easier to repeat things.[^repeating-blocks]

[^repeating-blocks]: If you want to repeat multiple lines of code, you just need to surround them
in curly braces `{ }`. So it isn't really that hard.
```{r, indent = "    ", fold = "so"}
do(3) * {
  hand1 <- sample(1:52, 3) 
  hand1
  hand1 <= 13
  sum(hand1 <= 13)
}
```

```{r}
sum(sample(1:52, 3) <= 13)
sum(sample(1:52, 3) <= 13)
sum(sample(1:52, 3) <= 13)
```

We just need to do that lots of times.

```{r}
set.seed(123456)
HandsOfThree <- do(10000) * sum(resample(1:52, 3) <= 13)
head(HandsOfThree) %>% pander()
```

Finally, we can use `tally()` to find the proportions we are looking for. 

```{r}
tally( ~ sum, data = HandsOfThree, format = "prop")
gf_dhistogram( ~ sum, data = HandsOfThree, binwidth = 1)
```

## A Statistical Question

The simulation method only provides an *estimate* of the probability.
If you recreate the examples abovce with a different random draw (either don't use `set.seed()` or 
use a different number in `set.seed()`), you will get somewhat different results.
We would like to know things like

* how much the estimates vary from simulation to simulation
* how many replications we should do in our simulations to get an estiamte with a desired
level of precision. (It seems pretty clear that more is better, but *how many more* do we need?)

These sorts of quetsion arise whenever we try to estimate something and form one of the big
ideas in statistics. Stay tuned for answers.

## Exercises

**1. Coins**  
a.	A fair coin is tossed 8 times. Use 10,000 simulations to estimate the probability distribution for the random variable X = number of heads.  
b.  	Using your result in (a), estimate the probability that at least 5 heads are produced.
c. 	Repeat (a) and (b) for a biased coin where the probability of a head is .66.

**2. Dice** 
Three fair dice are tossed.  Let X be the sum of the values produced.

a. Use 10,000 simulations to estimate the probability distribution of X.
b. Use your result in (a) to estimate the probability that the sum is less than 10.

**3. Softball** 
Carol has a batting average in softball of .440. (This means she gets a hit 44% of the time.)
If she has 5 at bats in a game, what is the probability that she gets exactly 2 hits?  What is the
probability that she gets at least 2 hits?  Simulate her performance in 10000
games to estimate these probabilities.

**4. Cards**
Use 10000 simulations to estimate the probability that in a five card hand, at least four of the cards 
are red (either a heart or a diamond).  Include the R-commands you used and the outputs.

