---
title: "Conditional Probability"
author: "Stat 241"
output:
  html_document:
    theme: yeti
    css: ../styles/ds303-notes.css
  df_document: default
  
---

\def\tor{\operatorname{or}}
\def\tand{\operatorname{and}}
\def\tnot{\operatorname{not}}
\def\Prob{\operatorname{P}}

<script src="../styles/folding.js"></script>

```{r hooks, message=FALSE, include = FALSE}
knitr::knit_hooks$set(fold = function(before, options, envir) {
  if (before) {
    return(glue::glue('<div fold = "{options$fold}">'))
  } else {
    return('</div>\n')
  }
})
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 5, 
  fig.height = 2,
  fig.align = "center"
)
library(mosaic)
library(pander)
theme_set(theme_bw())
set.seed(20210108)
```




$\Prob(A \mid B)$ answers the question: Of the times that $B$ happens, how often does $A$ also happen?
Common ways this is expressed include

* The probability of $A$ given $B$
* The probability of $A$ conditional on $B$
* The probability of $A$ if $B$ 
* The probability that $A$ happens when $B$ happens

When $\Prob(B) \neq 0$, then
$$
\Prob(A \mid B) = \frac{\Prob(A \tand B)}{\Prob(B)} = \frac{\Prob(\mbox{both})}{\Prob(\mbox{condition})}
$$

**Note:** Usually, $\Prob(A \mid B)$, $\Prob(B \mid A)$ and $\Prob(A \tand B)$ are all 
different.  It is critical to know which of these three applies in a given situation and to use
the notation correctly.


## Practice with Conditional Probability

A group of 10 men and 15 women were polled as to whether they preferred eating
an orange or eating an apple.  The results of the poll are given below

```{r, echo = FALSE}
tibble(sex = rep(c('male', 'female'), times = c(10, 15)),
       fruit = c(
         rep(c('apple', 'orange'), c(4, 6)),
         rep(c('apple', 'orange'), c(10, 5))
         )
       )  %>% 
  tally( ~ sex + fruit, data = .) %>%
  pander::pander()
```



Suppose we put the names of all the people into a hat and select one randomly.
Consider the following events.

 * $M$: The selected person is Male.
 * $A$: The selected person's prefers apples to oranges.
 
For each of the following, (a) express the probability in words, (b) determine the probability.
For conditional probabilities, compute the probabilities two different ways.

1. $\Prob(A)$
\vfill
2. $\Prob(\tnot {A})$
\vfill
3. $\Prob(M)$
\vfill
4. $\Prob(\tnot {M})$
\vfill
5. $\Prob(A \tand M)$
\vfill
6. $\Prob(A \mid M)$
\vfill
7. $\Prob(M \mid A)$
\vfill
8. $\Prob(M \mid \tnot{A})$
\vfill
9. $\Prob(\tnot{M} \mid A)$

\newpage

## Product Rule 

Applying a little algebra, we get the following product rule:

$$
\Prob(A \tand B) = \Prob(A) \cdot \Prob(B \mid A)
$$

## Independence

If $\Prob(A) \neq 0$ and $\Prob(B) \neq 0$ and $\Prob(B \mid A) = \Prob(B)$, then
we say that $A$ and $B$ are **independent events**.

Intuitively, A and B are independent when the probability that one of them occurs does
not depend on whether the other one occurred. 

## More Practice

1. Show that if $\Prob(B \mid A) = \Prob(B)$, then $\Prob(A \mid B) = \Prob(A)$.
 
2. The product rule is even simpler when $A$ and $B$ are independent.  What is it?
 
3. Show that 
$\Prob(A \tand B \tand C) = \Prob(A) \cdot \Prob(B \mid A) \cdot \Prob(C \mid A \tand B)$.
 
 
4. Analogous rules hold for intersections of more events as well.  Write down the rule for 
the intersection of 4 events.
 

5. What is the probability of rolling doubles (two numbers that match) with standard dice?
Do this two ways: (a) using the Equally Likely Rule and (b) using the Product Rule.
 


6. What is the probability of a five-card flush (all cards the same suit)?
Do this two ways: (a) using the Equally Likely Rule and (b) using the Product Rule.
 

<!-- 7. There are 18 students in our class. -->
<!-- What is the probability that two people in this class have the same  -->
<!-- birthday (month and day)? (Hint: Use the Complement Rule.) -->

<!--     What assumption must we make to do this calculation? Is that  -->
<!--     a reasonable assumption? -->

<!-- \vfill -->

7. If two 6-sided dice are rolled and the first one is a 5, what is the probability
 that the sum is 10?
 
 
8. If two 6-sided dice are rolled and at least one of them is a 5, what is the probability
 that the sum is 10?
 
 
10. If two 6-sided dice are rolled and the sum is 10, what is the probability that at least 
 one of them is a 5? 
 
11. A pair of fair dice is tossed.  Given that the sum of the dice is 7, what is
the probability that one of the dice came up 3?

12. A jar contains 5 red and 4 green chips.   Two chips are drawn without
replacement .  What is the probability that both are red?  What is the
probability that one is red and one is green?

13. If in the previous example, the drawing is with replacement, 
what are the probabilities of the events?

14.	Four cards are dealt from an ordinary deck of playing cards.  What is the
probability that they are all aces?

15.	You have two jars of chips.  Jar 1 contains 4 red and 4 blue chips.  Jar 2
contains 5 red and 3 blue chips.  You toss a fair coin.  If the coin toss is
heads, you draw a chip from Jar 1.  If it is tails, you draw from Jar 2.  
What is the probability the you draw a red chip?  What is the probability that if the
drawn chip is red, it came from Jar1 (the coin was heads)?


13. A new test has been developed  for a serious disease.  Like all such tests
it’s not perfect.  It has a false positive rate of 2%; i.e., if a person doesn’t
have the disease, there is a 2% chance that the test will be positive.  It also
has a false negative rate of 3%; i.e., if a person has the disease, there is a
3% chance that the test is negative.  It is a relatively uncommon disease,
affecting only 0.1% of the population.  How good is the test?  If you test
positive, what’s the probability that you have the disease?

	<!-- Basic Events:			D = has the disease	+ = tests positive -->

	<!-- Probability asked for:		P(D | +) -->

	<!-- Given information:		P(D) = 0.001	P(+ | not D) = 0.02	P(- | D) = 0.03) -->


14.	A fair die is tossed 3 times.  What is the probability that a 1 occurs on each toss?

15. A manufacturer claims that 99% of its product will still be functioning
after 3 years.  I buy 5 of the company’s product.  What is the probability that
all 5 of these items will still be functioning after 5 years?


<!-- ## Exercises  -->

<!-- 1.	Suppose the disease in the  example were not so rare.  Suppose it were -->
<!-- present in 1% of the population.  If you tested positive, what would be the -->
<!-- probability that you had the disease? -->

<!-- 2.	Suppose the disease in the  example were common.  Suppose it were present in -->
<!-- 25% of the population.  If you tested positive, what would be the probability -->
<!-- that you had the disease? -->

<!-- 3.	Suppose a coin is biased and has a 25% chance of coming up heads.  The coin is tossed twice. -->

<!--     a.	What is the sample space? -->

<!--     b.	Are the outcomes equally likely? -->

<!--     c.	Use a tree diagram to compute the probabilities of the outcomes. -->

<!--     d.	What is the probability of getting at least one head. -->

<!-- e.	What is the probability of getting two heads given that at least one head occurs? -->

<!-- 4.	Quality control samples were taken for parts being produced by two assembly lines.  A sample of 100 parts produced by assembly line 1 contained 2 defective parts. A sample of 200 parts produced by assembly line 2 contained 6 defective parts.  The table below summarizes the results. -->

<!--  	LINE 1	LINE 2 -->
<!-- Defective	2	6 -->
<!-- Non-defective	98	194 -->







<!-- a.	Given that a part was found to be defective, what is the probability it came from Line 1? -->

<!-- b.	Given that a part was found to be non-defective, what is the probability it came from Line 1? -->

<!-- 5.	A fair coin it tossed 5 times. -->

<!-- a.	How many different outcomes are there? -->

<!-- b.	Are these outcomes equally likely? -->

<!-- c.	What is the probability that all the tosses produced heads? -->

<!-- d.	What is the probability that at least one tall is produced? -->

<!-- e.	What is the probability that exactly one head is produced? -->







<!-- 6.	Three cards are dealt from an ordinary deck of playing cards. -->

<!-- a.	What is the probability that exactly one of the cards is a spade? -->
<!-- b.	Let X = number of spades dealt.  Find that probability function, p(x), for the random variable X. -->

<!-- x	0	1	2	3 -->
<!-- p(x)	 	 	 	  -->






<!-- 7.	A device is assembled from two primary parts.  The first type of part is defective 3% of the time and the second type is defective 4% of the time.  The device functions properly if and only if both of the parts are non-defective.   What is the probability that a device made from these two parts will function properly?  What are you assuming in making your calculation?  Is your assumption reasonable? -->

<!-- 8.	A biased die is tossed two times.  The probability for each face is given below. -->

<!-- Outcome	1	2	3	4	5	6 -->
<!-- Probability	.1	.3	.2	.2.	.1.	.1 -->

<!-- a.	What is the probability of getting two  1’s? -->
<!-- b.	What is the probability of getting the same result on both tosses? -->


