---
title: "Propagation of Uncertainty"
author: "Stat 241"
output:
  html_document:
    theme: yeti
    css: ['../styles/ds303-notes.css', '../styles/worksheets.css']
  df_document: default
  pdf_document: default
---

\def\tor{\operatorname{or}}
\def\tand{\operatorname{and}}
\def\tnot{\operatorname{not}}
\def\Prob{\operatorname{P}}
\def\E{\operatorname{E}}
\def\Var{\operatorname{Var}}
\def\SD{\operatorname{SD}}
\def\Binom{{\sf Binom}}
\def\Norm{{\sf Norm}}
\def\T{{\sf T}}
\def\Chisq{{\sf Chisq}}
\def\Unif{{\sf Unif}}
\def\Exp{{\sf Exp}}
\def\Beta{{\sf Beta}}
\def\Gamm{{\sf Gamma}}
\def\Weibull{{\sf Weibull}}
\def\Pois{{\sf Pois}}

<script src="../styles/folding.js"></script>

```{r hooks, message=FALSE, include = FALSE}
knitr::knit_hooks$set(fold = function(before, options, envir) {
  if (before) {
    return(glue::glue('<div fold = "{options$fold}">'))
  } else {
    return('</div>\n')
  }
})
```


```{r setup, include=FALSE}
onLine <- TRUE
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 5, 
  fig.height = 2,
  fig.align = "center"
)
library(mosaic)
library(pander)
library(mosaicCalc)
library(triangle)
library(patchwork)
theme_set(theme_bw())
set.seed(20210108)
options(digits = 3)
```
 


## Error and Uncertainty

* **error** = estimate - estimand = actual - estimated

  * We won't know this number because we would not be estimating if we knew the estimand
 
<br>

* **uncertainty** -- a description of the distribution of errors (over the random process of creating an estimate)

<br>

* **standard uncertainty** -- the standard deviation of the distribution of errors (over the random process of 
creating an estimate)

<br>

### Example: sample mean

$$
\overline X \approx \Norm(\mu , \frac{\sigma}{\sqrt{n}})\; ,
$$ 
so the distribution of errors is 

$$
\overline X - \mu \approx \Norm(0, \frac{\sigma}{\sqrt{n}})\; ,
$$ 
and the standard uncertainty for this estimator is 

$$
\mbox{standard uncertainty} =  \frac{\sigma}{\sqrt{n}} \approx \frac{s}{\sqrt{n}}\; .
$$

We will denote the standard uncertainty of a quantity $X$ as $u_X$.

### Example: Difference in means

Similarly, for the difference in two population means (estiamted by taking the difference in two sample means), we have

$$
\mbox{standard uncertainty}
= \sqrt{ \frac{\sigma_1^2}{{n_1}}  + \frac{\sigma_2^2}{{n_2}} }
\approx \sqrt{ \frac{s_1^2}{{n_1}}  + \frac{s_2^2}{{n_2}} }
$$

### Goal: Estimate standard uncertainty for more situations.

* The methods above worked because we have rules for propagation of means and variances for linear transformations
and linear combinations.

* We will learn an approximate method that works when things are approximately linear.

* These methods will estimate standard uncertainty (ie, standard deviation of the estimator = SE), but they 
do not tell us about the *shape* of the distribution of errors. In many situtations the distribution of errors
will be at least approximately normal, but in other situations it may be quite different.

## Transforamtions of one random variable

An estimator is just a random variable, so we will derive our methods in terms of random variables
and then apply them to estimators.

We already know that 

* $\E(aX+b) = a \E(X) + b$
* $\Var(aX+b) = a^2 \Var(X)$

This means that when $Y = f(X)$ and $f$ is a linear function ($f(x) = ax + b$), then we know
how to compute the mean and variance for $Y$ if we know the mean and variance for $X$.

From calculus, you may remember that we can approximate arbitrary functions using the tangent line
at a point.

<div class="explain" label = "Show linearization">
$$
f(x) \approx f(a) + f'(a) (x - a)
$$

This approximation will be better if

* $x$ is close to $a$
* $f(x)$ is not too curvy (ie, $|f''|$ is small)

"Better" and "small" are relative -- be sure to keep units and the range of reasonable values in mind.
</div>


Let's apply this to our random variables. Suppose we know the mean and variance of $X$. We want to estimate
the mean and variance of $Y = f(X)$.

<div class="explain" label = "Show approximation">
* let $a = \E(X) = \mu$ 

  * Most of the values of $X$ will be near $\E(X)$, so that is a good place to estimate from.
  * $\E(X - \mu) = \E(X) - \mu = 0$, so that's also handy.
  
<br>
  
* Expected Value

\begin{align*}
\E(Y) &= \E(f(X)) 
\\ &\approx \E(f(\mu) + f'(\mu) (x - \mu)) 
\\ &= f(\mu) + f'(\mu) \E(x - \mu) 
\\ &= f(\mu)
\end{align*}

  * so $\E(f(X)) \approx f(\E(X))$.
  
<br>

* Variance  

\begin{align*}
\Var(Y) & = \Var(f(\mu) + f'(\mu) (X - \mu)) 
\\ 
&= \Var(f'(\mu) (X - \mu))  
\\
&= f'(\mu)^2 \Var(X)
\end{align*}

* Standard deviation
$$
\SD(Y) = |f'(\mu)| \SD(X)
$$
</div>

### Application to uncertainty estimation

We can't quite use the approximation formula above because we don't know $\mu$. (It's an unknown paramter that we 
are estimating.)  But we can use our estimate in place of $\mu$. 
Since the standard uncertainty is just the standard deviation of an estimator, 

<div class = "boxed">
If 

$$Y = f(X) \;, $$ 

then

$$
u_Y \approx | f'(\hat x) | u_X
$$
</div>

where $\hat x$ is our estimate.

We are simply replacing the mean $\mu$ with our estimate $\hat x$. (Typically, $\E(X) = \mu$, so this estimate will be 
quite good if our sample size is large.)

<div class = "example">
**Example 1.**
Let $A$ be a square with side length $L$.
Suppose we estimate $L$ to be 3.7 with uncertainty $u_L = 0.2$ cm. (This is often written as $3.7 \pm 0.2$.  When you
see $\pm$ notation, you must also check to see how it is being used.)

What is our uncertainty in the computed area?

$A = L^2$, so our function is $f(x) = x^2$.  $f'(x) = 2x$.  So we get

\begin{align*}
u_A & \approx |f'(\hat L)| u_L 
\\ &= f'(3.7) \cdot 0.2 
\\ &= 7.4 \cdot 0.2 = 1.48
\end{align*}

We could express this as $`r 3.7^2` \pm 1.48$. (But stay tuned for a note about how many 
digits we should be reporting.)
</div>

<div class ="example">
**Example 1 (revisited)**. There is another way to estimate uncertainty of a transformed
measurement.  Let's just simulate. In this case, we simulate values of $L$ that are normally
distributed near 3.7 with a standard deviation of 0.2. Then we can compute a bunch of simulated
areas and see what the standard deviation is.

<div class = "explain" label = "Show simulations">

```{r}
L <- rnorm(10000, mean = 3.7, sd = 0.2)
A <- L^2
sd(~ A)
gf_histogram(~ A)
gf_qq(~A) %>% gf_qqline()
```

The estimate in this case is very close to the simulated value.

Compared to a normal distrubtion, this has somewhat "light tails" (the smallest values are not quite as small nor
then largest quite as large as we would expect). 

Seems to work pretty well in this case.
</div>
</div>


## Sig Figs and Uncertainty

The guidelines in this sections are based on commonly used practices in 
physics and engineering.

### What to record, What to report
 
When you record the results of a measurement for which there is an \emph{a priori}
estimate of uncertainty, the uncertainty should be recorded along with
the measurement itself.  Similarly, reports of quantities estimated from data
should also include estimated uncertainties.

As a general guideline, a properly reported scientific estimated quantity includes 
the following five elements: 


1. A number (the estimate)
2. Units (e.g., m  or kg or seconds) 
3. A statement about how it was measured or calculated 
4. A statement about most likely sources of (the largest components of) error
5. An estimate of the uncertainty

<div class = "example">
**Example**
If you measured the length of a pendulum using a meter stick, you
might report the measurement this way: 

* Length $= 0.834 \pm 0.002$ m 
* Measured with a meter stick from pivot point to the center of the steel weight. 
*	Uncertainty reflects the limited accuracy of measurement with a meter stick.

</div>
	
In plots, the number is given by the scales of the plot, the units are typically included
in the axes labels, uncertainties may be represented by ``error bars", and a statement 
describing the method of measurement or calculation should appear in the plot legend.
 
### How many decimal places?
 
Numerical values and their uncertainties should be recorded to the proper number of
decimal places.  Most software either reports too many significant digits or
rounds numbers too much.  For 
correct professional presentation of your data, follow these guidelines:

* **Rule 1:** The experimental uncertainty should be rounded to one significant
figure unless the leading digit is a 1, in which case, it is generally
better to use two digits. [Note: some people prefer that two digits be used
when the leading digit is a 2 as well.]

* **Rule 2:** A measurement should be displayed to the same number of *decimal places* 
as the uncertainty on that measurement.

Note carefully the difference between significant figure and decimal place.  


The following examples will help: 

<div class="example">
**Example**
The timer reports a value of 0.3451 seconds.  The uncertainty on
the measurement is 0.0038 seconds.    

* By Rule 1, the uncertainty should be
reported to one significant figure, so we round it to 0.004 seconds.   
* By Rule 2, the measurement must also be rounded to the third decimal place.  

Thus, the measurement should be reported as $0.345\pm0.004$ seconds. 
</div>
 
                                                 
<div class="example">
**Example**
The measured value is $7.92538 \cdot 10^4$, 
and its uncertainty is $2.3872 \cdot 10^2$.

* By Rule 1, the uncertainty should be rounded to one significant figure, so 
$2 \cdot 10^2$.  
* By Rule 2, we report the 
measurement to the same decimal place as the uncertainty, so $7.93 \cdot 10^4$.  

Putting it together, the measurement should be reported as $(7.93 \pm0.02) 10^4$.   

If we use the alternative method for Rule 2, we would report as
$(7.925\pm 0.024) 10^4$.   
</div>
 
                                                 
<div class="example">
**Example**
The estimated value is $89.231$, and its uncertainty is $0.1472$.    

* By Rule 1, 
the uncertainty should be rounded to two significant figures, so $0.15$.
* By Rule 2, we report the 
estimate to the same decimal place as the uncertainty, so $89.23  \pm  0.15$.  
</div>
 
                                                 

### Reporting numbers in a table
 
Multiple similar measurements should be reported in a table.  The column
headings should clearly and concisely indicate the quantity in each column; the
column heading must include the units.   Uncertainties should be listed in a
separate column, located just to the right of the measurement column.
(Sometimes, uncertainties are listed in parentheses after the estimate instead;
just make sure the header and legend of the table makes it clear what values are
being reported, and where.)

<div class = "example">
**Example**
A lab group calculated these numbers for kinetic energy and its uncertainty: 

------------------------------
 Kinetic Energy   uncertainty 
---------------- -------------
  0.8682              0.059 

  1.0661              0.071 

  1.0536              0.070 

  1.3881              0.058 
  
  0.8782              0.108 
-----------------------------

This should be reported with appropriate rounding (and units) as

------------------------------
 Kinetic Energy (J)   uncertainty 
--------------------- -------------
    0.87                 0.06 

    1.07                 0.07 

    1.05                 0.07 

    1.39                 0.06 

    0.88                 0.11 
-----------------------------

</div>

 


## Some Practice

**1.** Suppse you are estimating the area of a square. To do this, you measure the 
side length 30 times.  Your 30 measurements have a mean of 10.5 cm
and a standard deviation of 0.25 cm.
Use this to

a. Compute the standard uncertainty of the measured side length.
b. Estimate the standard uncertainy of the computed area.

**2.** 
10 measurements of the edge of a cube produce a sample mean $\overline{x} = 3.11$ in 
and a sample standard deviation $s = .13$ in. 
How should the volume of the cube be reported? 


<!-- 	SE(x ̅) = s/sqrt(10) = .04 -->

<!-- Volume(x ̅) = 3.113 ± SE(x ̅3) = 30.08 ± SE(x ̅3). -->

<!-- Since volume(x) = f(x) = x3 and f ‘(x) = 3x2,   -->

<!-- SE(x ̅3)   |3x ̅2|*SE(x ̅) = 29.02*.04 = 1.16.  So, the final report is -->

<!-- 		Volume = 30.08 ± 1.16 in3 -->

**3.** A temperature is reported as $357 \pm 2$ degrees Fahrentheit.
How should it be reported in Celsius? Do this two ways.

a. Use the approximation method we have just learned.
b. Work it out exactly using our rules for expected value and variance.
c. How do the two results compare? Why?

**4.**
An angle has size $\theta$, measured in radians.  The exact value of $\theta$ isn't known, 
but a sample of size $n = 20$ produces a sample mean of 1.21. and a sd of 0.09. 
How should the value of $\sin(\theta)$ be reported?


<!-- In this case f(x) = sin(x) and f ‘(x) = cos(x), so SE(sin(θ ̅  ))  |cos(θ ̅)|*SE(θ ̅). -->

<!-- SE(θ ̅) = 0.9 /sqrt(20) = 0.02,  cos(1.21) = .353, so SE(sin(θ ̅  ))  .01 -->

<!-- Since sin(1.21 ) = .94, the report would be -->

<!-- 	sin(θ) = 0.94 ± 0.01 -->

**5.**
An angle $\theta$ is reported as  $1.52 \pm .01$ (in radians),  
How should $\tan(\theta)$ be reported?

<!-- 	tan(θ) = tan(1.52) ± SE(tan(θ)) = 19.67 ± SE(tan(θ)).  -->

<!-- Since the dervative of tan is sec2, -->

<!--  SE(tan(θ))  |sec2(1.52)|*SE(θ) = 387.89*0.01 = 3.88.   -->

<!-- So, tan(θ) = 19.67 ± 3.88 -->

**6.**
	An angle is reported as $0.32 \pm .02$ radians. How should we report the cosine of the angle?

**7.**
The radius of a circle is reported as $r = 3.45 \pm .06$ in.  
How should we report the area of the circle?

**8.**
The radius of a sphere is reported as $r = 4.23 \pm .10$ in. 
How should we report the volume?


## Combinations of mutliple random variables

Coming soon.

## More Practice

**9.** Suppose $x ̂= 3.41 \pm .04$ is the estimate of the length of a rectangle and 
$y ̂= 2.34 \pm .02$ is the estimate of its width.  How do we report the area?

<!-- Area = f(x,y) = xy.  fx(x,y) = y, fy(x,y) = x -->

<!-- 	Area =  f(x ̂, y ̂) ± SE(f(x ̂, y ̂)) = 3.41*2.34 ± SE(f(x ̂, y ̂)) = 7.98 ± SE(f(x ̂, y ̂)) -->

<!--      SE(f(x ̂, y ̂))   √(〖f_x (x ̂,y ̂)〗^2 〖SE((x)) ̂〗^2+〖f_y (x ̂,y ̂)〗^2 〖SE(y ̂)〗^2 ) -->

<!-- 			= √(〖(y ̂)〗^2 〖SE(x ̂)〗^2+〖(x ̂)〗^2 〖SE(y ̂)〗^2 ) -->

<!-- 			√(〖2.34〗^2 〖*.04〗^2+〖3.41〗^2 〖*.02〗^2 )	  -->

<!-- 			= .12 -->

<!-- The final result:  Area = 7.98 ± 0.12 -->

**10.** The length  of a rectangle is reported as 
$\hat x = 3.05 \pm .02$ and the width of the rectangle is reported as 
$\hat y  = 5.45 \pm .11$.  How should perimeter of the rectangle be reported?

<!-- Perimeter = f(x,y) = 2x + 2y,  fx (x,y) = 2, fy(x,y) = 2 -->

<!-- Perimeter = f(x ̂, y ̂) ± SE(f(x ̂, y ̂)) = 2*3.05+2*5.45 ± SE(f(x ̂, y ̂)) = 17.00 ± SE(f(x ̂, y ̂)) -->

<!--      SE(f(x ̂, y ̂))   √(〖f_x (x ̂,y ̂)〗^2 〖SE((x)) ̂〗^2+〖f_y (x ̂,y ̂)〗^2 〖SE(y ̂)〗^2 ) -->

<!-- 			= √(2^2 〖*.02〗^2+2^2*〖.11〗^2 ) -->

<!-- 			= .22 -->

<!-- The final result:  Perimeter = 17.00 ± .22. -->


