---
title: "Random Variables"
author: "Stat 241"
output:
  html_document:
    theme: yeti
    css: ../styles/ds303-notes.css
  df_document: default
  pdf_document: default
---

\def\tor{\operatorname{or}}
\def\tand{\operatorname{and}}
\def\tnot{\operatorname{not}}
\def\Prob{\operatorname{P}}
\def\Binom{{\sf Binom}}
\def\Norm{{\sf Norm}}
\def\Unif{{\sf Unif}}

<script src="../styles/folding.js"></script>

```{r hooks, message=FALSE, include = FALSE}
knitr::knit_hooks$set(fold = function(before, options, envir) {
  if (before) {
    return(glue::glue('<div fold = "{options$fold}">'))
  } else {
    return('</div>\n')
  }
})
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 5, 
  fig.height = 2,
  fig.align = "center"
)
library(mosaic)
library(pander)
theme_set(theme_bw())
set.seed(20210108)
```



## Discrete Random Variable

A random variable $X$ is a **discrete random variable** if the range of $X$ is
finite or an infinite sequence of values. A discrete random variable can be
characterized by a function called a **probability mass function (pmf)** which
specifies the probability for each possible numerical value of the random
variable.
 	
$$
 	p(x) = \Prob(X = x)
$$
This function can be specified with a formula or in a table of probabilities.
Given the pmf we can calculate
the probability any event by adding the probabilities
for all of the these values in the event.

**Example 1.**
Toss a fair coin 5 times.  Let $X$ = number of heads produced. 

a. Create a probability table for $X$.
b. What is $\Prob(X \mbox{ is even})$?
c. What is $\Prob(X \ge 2)$?

### Binomial Random Variables

This example is an instance of an important kind of discrete random variabl called a 
**binomial random variable**. Binomial random variables arise in situations which 
have the following properties.

1. The random process consists of a **predetermined number of trials** (usually denoted $n$).
2. Each trial has **two outcomes** (generically called success and failure).
3. The **probability** of success **is the same for each trial** (often denoted $p$).
4. Each trial is **independent** of the others.
5. The random variable counts the **number of successes** in the $n$ trials.

Such a variable is dennoted by $X \sim \Binom(n, p)$.

There is a handy formula for the pmf of a $\Binom(n, p)$ random variable:

$$
p(x) = \Prob(X = x) = \binom{n}{x} p^x (1-p)^{n-x}
$$

where $\binom{n}{x} = \frac{n!}{x! (n-x)!}$.  $\binom{n}{x}$ is read "n choose x" and counts the
number of ways to pick a set of $x$ items from a collection of size $n$.
This can be computed in R using the `choose()` function.

```{r}
choose(5, 2)
factorial(5) / (factorial(2) * factorial(3))
```

**Example 2.** If we roll a fair die 10 times, what is the probability of getting 2 or more sixes?

\vfill

\newpage 

**Example 3.** 
Two cards are dealt from an ordinary deck of playing cards.  Let $X$ = number of aces dealt.

a. Create a probability table for $X$.
b. Is $X$ a binomial random variable?  If so, what are $n$ and $p$? If not, why not?
c. What is $\Prob(X \ge 1)$? 

\vfill
\vfill

**Example 4.**
A fair coin is tossed until a head is produced.  Let $X$ = number of tosses.

a. Create a probability table for $X$.
b. Can you give a formula for the pmf for $X$?
c. Can you show that the sum of all the probabilities is 1?
d. What is $\Prob(X \ge 4)$?
e. What is the probability it takes an even number of tosses to production first head? ($\Prob(X \mbox{ is even}$)

\vfill
\vfill

\newpage


## Continuous Random Variables

A **continuous random variable** is a random variable whose range is an interval of real numbers. 

**Examples**

* Randomly select an adult male:  Let $X$ = (exact) height in inches

* Randomly select a washer produced by a particular machine.  Let *X* = (exact) diameter in mm.
	
* Randomly select a battery and use it to power a device. 
Let $X$ = (exact) time until battery can no longer power device.


### Density histograms and density plots of data

Let's consider the heights of the adult children in the `Galton` data set. Here is a **density histogram**.

```{r}
gf_dhistogram( ~ height, data = Galton)
```

The density scale is chosen so that ___________________________________________.

Now let's look at a density plot.

```{r}
gf_dhistogram( ~ height, data = Galton) %>%
gf_dens( ~ height, data = Galton, size = 1.2)
```

This provides a "smooth" version of the histogram and also has the property that 

\vfill

\newpage

### Density Curves and Density Functions

A continuous random variable is described by a **probability density function (pdf)**.
The plot of a pdf will look just like curve in a density plot.

<!-- For a continuous random variable $X$, if $x$ is in the range of $X$, $\Prob(X = x) = 0$,  -->
<!-- so characterizing the probability of $X$ by the function $p(x) = \Prob(X = x )$ doesn’t work.  -->
<!-- Instead, the probability characterization is its **probability density function**. -->
<!-- , f(x).   -->

Probability density functions always have two important properties:

\bigskip

1. <!-- non-negative -->

\bigskip

2. <!-- total area = 1 -->

\bigskip

We determine probabilities from a pdf but taking the area under
the curve over the region corresponding to our event (ie, by integration).

$$
\Prob( a \le X \le b) = \int_a^b f(x) \; dx
$$


**Example 5.**
Let $X$ be a number randomly chosen from the interval $[0, 2]$ in such a way that all numbers are equally likely.
We call $[0, 2]$ the **support** of $X$ is [0,2].  
Since no value of $X$ is more likely to be selected than any other value, the density function $f(x)$ 
must be a constant on [0,2].  

a. What is the constant value?

b. What is $\Prob(X = 1)$?

b. What is $\Prob(0 \le X \le 1)$?

c. Waht is $\Prob(1 \le X \le 3/2)$?

d. What is $\Prob(1 \le X \le 3)$?


\vfill
\vfill
\vfill


A random variable $X$ whose pdf is constant (where it is non-zero) is said to have a **uniform distribution**.
We will denote this as  $X \sim \Unif(a, b)$, where $a$ and $b$ are the upper and lower limits of the support.

\newpage


**Example 6.** Let $f(x) = x/2$ for $x \in (0,2)$ (and 0 elsewhere).  We can write this as 
$f(x) = x/2 \cdot [\![x \in (0,2) ]\!]$ or $f(x) = x/2$ on $(0,2)$.

a. Verify that $f(x)$ is a probability density function.

b. Compute $\Prob(1 \le X \le 3/2)$.

b. Compute $\Prob(1 \le X \le 4)$.

\vfill
\vfill


### The Cumulative Distribution Function (cdf) for a continuous random variable 

The **cdf** for $X$ is defined by

$$
F(x) = \Prob(X \le x)
$$

**Example 7.**
Let $X$ have a uniform distribution on $[0,4]$. 

a. pdf: $f(x) =$

\vfill


b. cdf: $F(x) =$

\vfill



**Example 8.** 
Let $X$ be the random variable whose pdf is $f(x) = x/2 \cdot [\![ x \in (0,2) ]\!]$. 
Find the cdf for $X$.

\vfill
\vfill


\newpage

### Using the cdf to compute probabilities

If $F(x)$ is the cdf for the random variable $X$, then $\Prob(a \le X \le b) =$  ________________________


**Example 9.** Let $F(x)= x^2$ on $[0,1]$

a. What is $F(x)$ when $x < 0$?

a. What is $F(x)$ when $x > 0$?

a. What is $P(X \le 1/2)$?

a. What is $P(1/2 \le  X \le 3/4)$?

a. What is $P(-2 \le X \le 1/2)$?

a. What is the pdf for $X$?

\vfill
\vfill
\vfill


<!-- Using the cdf to compute the pdf -->

<!-- 	If f(x) is the pdf and F(x) is the cdf, then f(x) = F’(x). -->



**Example 10.** Let $f(x) = e^{-x}$ on $(0,\infty)$ (and 0 elsewhere).

a. Show that $f$ is a pdf.
b. Let $X$ be the corresponding random variable. What is $\Prob(X > 2)$?
c. What is the cdf for $X$?
d. Use the cdf to find $\Prob(X < 1)$.

\vfill
\vfill
\vfill

\newpage

## More Practice 

**1.**
Parts coming off an assembly line have a 1% chance of being defective.  If 3
parts are randomly chosen from this line and X is the number of defective parts.

a. Compute the probability function p(x) for X.  

b. What is the probability that at least one of the three is defective?

**2.** Parts coming off an assembly line have a 1% chance of being defective.
All of the parts coming off the line are inspected. 
Let $X$ be the number parts inspected up to and including the first defective part.

a. Is $X$ continuous or discrete?

b. What is the support of $X$?

a. Find the probability mass function $p(x)$ for $X$.  

a. What is the probability that the first defective part is the 100th part?

**3.**
A biased coin has a 40% chance of producing a head.  If it is tossed 10 times,

a. What is the probability of getting exactly 3 heads?

a. What is the probability of getting 3 or more heads. 
(This can be calculated in two different ways. The easier way uses the complement rule.)

**4.** 
a. Find the value of $C$ for which the function $f(x) =  C x^2 \cdot [\![ x \in (0,2) [\!]$ 
is a pdf.

b. Use the pdf to find $\Prob(0 \le X \le 1)$ and $\Prob(1 \le X \le 5)$.

c. Use the pdf to compute the cumulative distribution function $F(x)$.

**5.**
a. Find the value of $C$ for which the function $f(x) = \frac{C}{x^2} \cdot [\![ x \ge 1 ]\!]$ is a pdf.

b. What is $\Prob(X \le 2)$?

c. What is $\Prob(X > 3)$?

d. Find the cumulative distribution function $F(x)$.

e. Use the cumulative distribution function to find $\Prob(2 \le X \le 5)$.
